# Percolation theory

## Dependencies

```{r}
# Load required packages
library(igraph)    # for network analysis
library(ggplot2)
library(ggraph)
library(patchwork)
library(tidygraph)
library(RColorBrewer)
```

## Introduction


The term percolation normally refers to the process whereby a fluid moves slowly through a porous material, for example, the percolation of rain water through rocks gives rise to aquifers. The study and modelling of this infiltration process in physics, chemistry and material science is known as percolation theory. If you think about it, a rock can be modelled as a three-dimensional lattice (if the rock was cube-shaped, this lattice would be like a Rubik cube but instead of being $3 \times 3\times 3$, it is $n\times n \times n$). Then, the pores in the rock would be represented by "open sites" in this lattice (this would correspond to missing pieces in the Rubik cube) and the bits of the rock where there is material and therefore, no chance for the water to go through, would be represented by "closed sites" (this would correspond to pieces in the Rubik cube that are actually present). A typical question in percolation theory would then be as follows. If sites may be open with probability $p$ or closed with probability $1-p$ and these probabilities are assumed to be independent for each site, what is the probability that the water can go through the top to the bottom of the rock for a given value of $p$?

In recent decades, the mathematical study of percolation has been applied in a more general way, to understand the behavior of connected clusters or pathways in any system that can be modelled as a lattice or as a network. While the formulation of the percolation models varies slightly from lattices to networks, it follows the same principles. For example, let's consider the spread of a certain pandemic, where contagion between two humans may take place with probability $p$ if an infected individual spends more than 10 minutes at less than 2 meters away from a healthy one. Then, we can think of the pandemic as a fluid moving slowly, but instead of going through a porous rock, it goes through a social network of people. The people can be modelled as the nodes of the social network, and edges are present between two individuals if they spend long enough in close physical contact. Then, each edge or connection has an associated probability of contagion $p$. Applying percolation theory, we could answer the question of "what are the chances that the pandemic makes its way from city A to city B given the structure of the network and the fact that the probability of contagion is $p$?"

Percolation theory can therefore be used in a variety of contexts to better understand the properties of a system characterised by the connections between its components. Here, we use percolation models to improve our understanding of the structure and resilience of the African road network.

As before, we start by loading the data corresonding to the nodes and edges of the African road network:

```{r}
# Read edge and node data from URLs
df_edges <- read.csv("https://raw.githubusercontent.com/rafaelprietocuriel/AfricanUrbanNetwork/main/AfricaNetworkEdges.csv")
df_nodes <- read.csv("https://raw.githubusercontent.com/rafaelprietocuriel/AfricanUrbanNetwork/main/AfricaNetworkNodes.csv", encoding='UTF-8')
```


```{r}
# Filter the data frame df_nodes to only include rows where the 'Region' column is equal to "South"
df_nodes <- subset(df_nodes, Region == "South")
```

```{r}
# Filter the data frame df_edges to only include rows where the 'from' column is in the list of Agglomeration_ID values from df_nodes
df_edges <- subset(df_edges, from %in% df_nodes$Agglomeration_ID)

# Filter the data frame df_edges to only include rows where the 'to' column is in the list of Agglomeration_ID values from df_nodes
df_edges <- subset(df_edges, to %in% df_nodes$Agglomeration_ID)
```

And we create an undirected graph from these data frames:

```{r}
# Create an undirected graph 'G' from a data frame 'df_edges' representing edges and a data frame 'df_nodes' representing vertices
G <- graph_from_data_frame(d = df_edges,
                           vertices = df_nodes,
                           directed = FALSE)
```

As before, we can visualise this network by running the code below

```{r}
# Assign a 'size' attribute to vertices in graph 'G' based on a function of population data
V(G)$size <- 0.5*(V(G)$Pop2015/10000)^0.4

# Plot the graph 'G' with specific visual attributes
plot(G, 
     vertex.size = V(G)$size,          # Set vertex size based on the 'size' attribute
     edge.arrow.size = 0.15,           # Set arrow size for directed edges
     edge.arrow.width = 0.2,          # Set arrow width for directed edges
     edge.curved = 0.1,               # Set edge curvature
     edge.width = 1,                  # Set edge width
     edge.color = "gray90",           # Set edge color
     vertex.color = "red",            # Set vertex color
     vertex.frame.color = "black",    # Set vertex frame (border) color
     vertex.frame.width = 0.2,        # Set vertex frame (border) width
     vertex.label = " ",              # Set vertex labels to empty
     vertex.label.color = "black",    # Set vertex label color
     vertex.label.cex = 0.65)         # Set vertex label size 
```

## Modelling community formation in the African road network

At the end of the previous session, we used the fast greedy algorithm to detect communities based on the topology of the African road network. We used an R implementation of the fast greedy algorithm through the built-in function `cluter_fast_greedy()`. While a built-in function gives us results in a very straightforward way, it also makes the process less transparent. In this section, we implement a percolation theory approach to model the formation of communities of nodes (cities) as the connectivity properties of the network are varied. 


```{r}

# Create a list to store the plots
plots_list <- list()
i=1

# Iterate over thresholds
for (j in seq(60, 360, 60)) {
  
  df_edges_perco <- subset(df_edges, timeUCB < j)
  
  G_perco <- graph_from_data_frame(d = df_edges_perco,
                           vertices = df_nodes,
                           directed = FALSE)
  
  # Find connected components
  component_info <- components(G_perco)

  # Define a color palette for connected components (replace with your own colors)
  component_colors <- rainbow(max(component_info$membership) + 1)

  # Create a data frame with node names and corresponding colors
  node_colors <- data.frame(
    name = df_nodes$Agglomeration_ID,
    color = component_colors[component_info$membership + 1]
  )
  
  # For isolated nodes, assign gray color
  isolated_nodes <- V(G_perco)[degree(G_perco) == 0]
  # Assuming 'attribute_name' is the attribute you want to retrieve
  attribute_values <- get.vertex.attribute(G_perco, "AgglomerationID")[isolated_nodes]

  
  node_colors$color[node_colors$name %in% attribute_values] <- "gray"
  
  # Assign a 'size' attribute to vertices in graph 'G' based on a function of population data
  V(G_perco)$size <- 0.5*(V(G_perco)$Pop2015/10000)^0.4

  custom_layout <- data.frame(
  name = V(G_perco)$name,  # Node names from the graph
  x = df_nodes$x,  # Custom x-coordinates
  y = df_nodes$y   # Custom y-coordinates
  )
  
  # Plot the graph 'G_' with specific visual attributes
  plot <- ggraph(as_tbl_graph(G_perco), custom_layout) +
    geom_edge_link() +
    geom_node_point(aes(color = node_colors$color), 
                    size = V(G_perco)$size) +
                    theme(legend.position = "none")
    
     # plot(G_perco, 
     # vertex.size = V(G_perco)$size,         # Set vertex size based on the 'size' attribute
     # edge.arrow.size = 0.15,          # Set arrow size for directed edges
     # edge.arrow.width = 0.2,          # Set arrow width for directed edges
     # edge.curved = 0.1,               # Set edge curvature
     # edge.width = 1,                  # Set edge width
     # edge.color = "gray90",           # Set edge color
     # vertex.color = colors[component_membership],            # Set vertex color
     # vertex.frame.color = "black",    # Set vertex frame (border) color
     # vertex.frame.width = 0.2,        # Set vertex frame (border) width
     # vertex.label = " ",              # Set vertex labels to empty
     # vertex.label.color = "black",    # Set vertex label color
     # vertex.label.cex = 0.65)         # Set vertex label size 
  
  # Add the plot to the list
  plots_list[[i]] <- plot
  
  i <- i+1
}
```



```{r}
plots_list %>%
  wrap_plots(nrow = 2) 

```








## Measuring network robustness



```{r}
# Create an empty data frame to store component information
components <- data.frame(nodeID = integer(0), component = integer(0), threshold = integer(0), gcc = integer(0))

# Create empty lists to store thresholds, gccs, and ncs
thresholds <- list()
gccs <- list()
ncs <- list()
```



## Original by removing

```{r}

# Iterate over thresholds
for (i in seq(0, max(df_edges$l))) {
  # Create a copy of the graph G
  G_ <- G 
  
  # Find indices of edges with lengths greater than the current threshold (i)
  edges_to_remove <- which(E(G_)$l > i)
  
  # Delete edges from G_ based on their indices
  G_ <- delete_edges(G_, edges_to_remove)
  
  # Get connected components of the modified graph G_
  connected_components <- components(G_)
  
  # Create a data frame 'df_threshold' containing node IDs, component indices,
  # threshold values, and sizes of connected components
  df_threshold <- data.frame(nodeID = df_nodes$Agglomeration_ID, component = connected_components$membership, threshold = rep(i, times= nrow(df_nodes)), gcc = rep(max(connected_components$csize), times= nrow(df_nodes)))
  
  # Append 'df_threshold' to the 'components' data frame
  components <- rbind(components, df_threshold)
  
  # Append the current threshold value to the 'thresholds' list
  thresholds <- append(thresholds, i)
  
  # Append the maximum connected component size to the 'gccs' list
  gccs <- append(gccs, max(connected_components$csize))
  
  # Append the number of connected components to the 'ncs' list
  ncs <- append(ncs, connected_components$no)
}

# Display the first few rows of the 'components' data frame
head(components)
```



```{r}
# Plot the threshold values on the x-axis and the maximum connected component sizes (gccs) on the y-axis
plot(thresholds, gccs)
```

```{r}
# Plot the threshold values on the x-axis and the number of connected components (ncs) on the y-axis
plot(thresholds, ncs)
```


